\textcolor{blue}{Local Maximum Entropy Material Point Method applied to quasi-brittle fracture.}

### 2020-11-04 2020_EFM_MPM_Eigensoftening.tex
Updates from Overleaf

\newacronym{mpm}{MPM}{Material Point Method}
\newacronym{otm}{OTM}{Optimal Transportation Meshfree}
\newacronym{sph}{SPH}{Smoothed Particle Hydrodynamics}
\newacronym{fem}{FEM}{Finite Element Method}
\newacronym{xfem}{XFEM}{Extended Finite Element Method}
\newacronym{efgm}{EFGM}{Element-Free Galerkin Method}
\newacronym{lme}{LME}{Local Maximum-Entropy}
\newacronym{flip}{FLIP}{Fluid Implicit Particle}
\newacronym{gimp}{GIMP}{Generalized Interpolation Material Point}
\newacronym{igimp}{iGIMP}{Implicit GIMP}
\newacronym{ugimp}{uGIMP}{Uniform GIMP}
\newacronym{ddmp}{DDMP}{Dual Domain Material Point}
\newacronym{ctls}{CTLS}{Conservative Taylor Least Squares}
\newacronym{npc}{NPC}{Newmark Predictor-Corrector}
\newacronym{fe}{FE}{Forward Euler}
\newacronym{cramp}{CRAMP}{CRAcks with Material Points}

@@

\newglossaryentry{dirichlet-boundary}
{
  name=$ \Gamma_d$,
  description={Essential or Dirichlet boundary conditions over $\partial\varOmega$}
}

\newglossaryentry{neumann-boundary}
{
  name=$ \Gamma_n$,
  description={Natural or Nemann boundary conditions over $\partial\varOmega$}
}

@@

@@

\newglossaryentry{Ceps}
{
  name=$C_{\epsilon}$,
  description={Normalizing constant}
}

\newglossaryentry{heps}
{
  name=$h_{\epsilon}$,
  description={Band width}
}

@@

@@

\newglossaryentry{gamma-lme}
{
  name=$\widehat{\gamma}$,
  description={Dimensionless parameter to control the value of the
    thermalization parameter $\beta$ of the \acrshort{lme} shape functions.}
}

@@

\journal{Engineering Fracture Mechanics}
% \journal{Computers and Structures}

@@

\textcolor{blue}{The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian method where the continuum \gls{domain} is discretised with a finite set of material points (also known as particles in this research) $\hat{\Omega} = \{ \vec{x}_p, p \in \mathcal{C} \} \subset \Omega$, where $\mathcal{C} = 1, \ldots, n_p$. Any particle field such as position, velocity, mass, volume and stress denoted by $\vec{x}_p$, $\vec{v}_p$, $m_p$, $V_p$ and $\tens{\sigma}_p$, respectively, are assigned to each material point.  In addition to this set of material points, a set of fixed background nodes is introduced to compute the balance of momentum equation. This set of nodes is defined as  $\textbf{X} = \{ \vec{x}_I, I \in \mathcal{B} \} \subset \mathbb{R}^{2}$, where $\mathcal{B} = 1, \ldots, n_I$. In the present research, any nodal quantity is denoted by ($\square_I$).}

@@

function width is defined as $U(\vec{x}_p,N_I) \coloneqq
\sum_I N_I |\vec{x}_p - \vec{x}_I |^2$, and $\beta$ is a
regularisation o \textit{thermalization} parameter such that for

@@

(0,\infty)$. Notice that $\beta$ has units of $[L]^{-2}$,

@@

enough efficient method. For an uniform nodal spacing, $\beta$ can be

@@

\subsection{Fracture modelling approach}
\label{sec:2.3}
Within the \textcolor{red}{\acrshort{mpm} formulation context}, fracture can be
modelled by failing particles according to a suitable failure
criterion. When material points are failed, they are assumed to have
null stress tensor. To reproduce this behaviour in the present
research, the eigensoftening algorithm is introduced in the 
\acrshort{mpm} framework as an alternative approach to the decohesion
model \cite{Schreyer_2002}. The eigensoftening concept
was \textcolor{red}{developed initially} by Navas {\it et al.}
(2017)\cite{Navas_2017_ES} as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. The
  evaluation of both approaches \cite{Navas_2017_ES} against
  experimental measures show that the eigenerosion algorithm significantly overestimates the tensile stress and the strain peaks, Meanwhile eigensoftening captures the forces and crack
patterns accurately. Furthermore, the accuracy of this algorithm has been proven for \textcolor{red}{mixed-mode} fracture patterns such as the ones obtained with fiber \textcolor{red}{reinforced} concrete (FRC),
\cite{Navas_2018_ES}.\\

The key idea behind the eigenerosion algorithm is the computation of the
energy-release rate attendant to the failure of material point $p$,
\begin{align}
  \label{eq:energy-release-EE}
&G_p^{k+1} = \frac{C_{\epsilon}}{m_p^{k+1}}  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q W_q^{k+1}\\
  \label{eq:mass-EE}
&m_p^{k+1} =  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q  
\end{align}
where $B_{\epsilon}(x_p^{k+1})$ is a $^n$dimensional sphere of radius $\epsilon$
centered at $x_p^{k+1}$. The particles which lie under this sphere are
known as the $\epsilon$-neighborhood of the material point (see
\cite{Pandolfi_2012}). This concept is conveniently sketched in
Figure \ref{fig:Failed-particles}.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure-Particle-failed}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical
    opening displacement.}
  \label{fig:Failed-particles}
\end{figure}
%%%%%%%%%%%
Other parameters are the mass of the neighborhood $m_p^{k+1}$, the
current free-energy density per unit mass  $W_q^{k+1}$ and the
normalizing constant $C_{\epsilon}$, which also defines the
  $\epsilon$-neighborhood size, being $\epsilon = C_\epsilon h$, and
  $h$ is a measure of the distance between nodes.

The failure criterion \textcolor{blue}{considers that a material point fails}
when $G_p^{k+1}$ surpasses the critical energy release rate that
measures the material-specific energy, $G_F$. The convergence of this
approach has been analyzed by Schmidt {\it et al.}
(2009)\cite{Schmidt_2009}, who proved that $G_F$ converges to the Griffith
fracture when the discretisation size tends to zero. It is necessary to
point out that, when a material point overpass the critical energy, its
contribution to the internal forces vector is set to zero. \textcolor{red}{However,} its
contribution to the mass matrix is preserved.\\

@@

the static fracture energy per unit of area, $G_F$. Notice how a
cohesive crack appears when the maximum tensile strength, $f_t$, is
reached. Once the crack, $w_n$, reaches the value of the critical
crack opening, $w_c$, a stress-free crack is attained.

@@

    $G_f$, $f_t$ is the tensile strength, and $w_c$ is the critical

@@

$\epsilon$-neighborhood of the material point $\vec{x}_p^{k+1}$ as

@@

material point in the $\epsilon$-neighborhood. At this point, an \textit{effective} strain $\varepsilon_q$ is introduced in order to obtain the variation of

@@

\delta\varepsilon_q$. Now, with the assumption that the effective

@@

  \frac{\delta \tens{\epsilon}_p}{m_p} \sum_{x_q^{k+1} \in

@@

 \tens{\sigma}_{\epsilon,p}\ \delta\varepsilon_p$. The softening behaviour is

@@

tensile strength, $f_t$. This consists of a reduction of the internal

@@

\cite{Navas_2017_ES} introduced a band width parameter $h_{\epsilon}$ in order to relate the crack opening with the strain.

@@

opening $w_n$ within the band width $h_{\epsilon}$. Therefore, 

@@

\begin{table}
  \centering
  \begin{tabular}[]{c c c c c c c}
    \hline
      &   $\rho$   & $\text{f}_c$ & $\text{f}_t$ & $G_F$ &   E   & $\text{d}_{max}$ \\
      & (kg/m$^3$) &     (MPa)    &     (MPa)    & (N/m) & (GPa) & (mm) \\
    \hline
Value &    2368    &     102.7    &      5.4     &  141  &   31  &  12 \\
    \hline
  \end{tabular}
  \caption[Mechanical properties of thje concrete]{Mechanical
    properties of the high strenght concrete.}
  \label{tab:concrete-properties}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the present research, a 2D setting of the experiment has been
adopted. Both the hammer and the concrete beam are explicitly
represented in order to properly capture the impact. Several levels of
discretisation were required to assess the objectiveness of the
obtained results. The results here exposed are from a discretisation
of 21576 material points and 3943 nodes. To achieve better results, an
unstructured grid layout has been adopted focusing the minimum nodal
spacing, 1.2 mm, in the middle of the beam and in the impact surface
of the hammer. For this simulation, \gls{gamma-lme} was fixed to
1.8, this low value of the regularization parameter smooth high stress
overshots after the impact. Here the stabilization properties of the \acrshort{lme}
described in \cite{Molinos2020} becomes crucial to achieve
non-spurious results. Standard linear element were tested
with unsatisfactory results since grid-crossing occurs in the hammer
head producing non realistic forces.\\

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \KwInput{For each $p$, $\epsilon$-neighbourhood, $f_{t,p}$, $h_{\epsilon,p}$, $w_c$}
  \KwOutput{Return damage parameter $\chi = \{\chi_p\}$}
  \KwData{Testing set $x$}
  $\chi_p \leftarrow \chi_p^{k}$ \;
  \For{$p$ to $N_p$}
  {
    \If{$\chi_p = 0$ \textbf{and} $\epsilon_{f,p} = 0$}
    {
      \For{$q \in B_{\epsilon,p}$}
      {
        $\sigma_{q,I} \leftarrow \text{getEigenvaluesOf} (\sigma_{q}) $ \;
        \If{$\chi_q < 1$}
        {
          $ \sum m_p\sigma_{p,I} \leftarrow \sum m_p\sigma_{p,I} + m_q\sigma_{q,I}$ \;
        }
        $ m_p \leftarrow m_p + m_q$ \;
      }
      $\sigma_{p,\epsilon} \leftarrow \frac{1}{m_p} \sum m_p\sigma_{p,I}$\;
      \If{$\sigma_{p,\epsilon} > f_{t,p}$}
      {
        $\epsilon_{f,p} = \epsilon_{I,p}$ \;  
      }
      \ElseIf{$\chi_p \neq 1$  \textbf{and} $\epsilon_{f,p} > 0$}
      {
        $\chi_p^{k+1} \leftarrow \min\Big \{1 , \max \{\chi_p^{k},
        \frac{(\epsilon_{I,p}- \epsilon_{f,p})\ h_{\epsilon,p}}{w_c}
        \} \Big \}$ \;
      }
    }
  }
    \label{alg-eigens}
    \caption{Compute damage parameter $\chi$}
\end{algorithm}

>>>>

\newacronym{mpm}{MPM}{Material Point Method}
\newacronym{otm}{OTM}{Optimal Transportation Meshfree}
\newacronym{sph}{SPH}{Smoothed Particle Hydrodynamics}
\newacronym{fem}{FEM}{Finite Element Method}
\newacronym{xfem}{XFEM}{Extended Finite Element Method}
\newacronym{efgm}{EFGM}{Element-Free Galerkin Method}
\newacronym{lme}{LME}{Local Maximum-Entropy}
\newacronym{flip}{FLIP}{Fluid Implicit Particle}
\newacronym{gimp}{GIMP}{Generalized Interpolation Material Point}
\newacronym{igimp}{iGIMP}{Implicit GIMP}
\newacronym{ugimp}{uGIMP}{Uniform GIMP}
\newacronym{ddmp}{DDMP}{Dual Domain Material Point}
\newacronym{ctls}{CTLS}{Conservative Taylor Least Squares}
\newacronym{npc}{NPC}{Newmark Predictor-Corrector}
\newacronym{fe}{FE}{Forward Euler}
\newacronym{cramp}{CRAMP}{CRAcks with Material Points}
\newacronym{frc}{FRC}{Fiber Reinforced Concrete}

@@

\newglossaryentry{dirichlet-boundary}
{
  name=$\Gamma_d$,
  description={Essential or Dirichlet boundary conditions over $\partial\varOmega$}
}

\newglossaryentry{neumann-boundary}
{
  name=$\Gamma_n$,
  description={Natural or Nemann boundary conditions over $\partial\varOmega$}
}

\newglossaryentry{m}
{
  name=$m$,
  description={Describes the scalar mass field}
}

@@

\newglossaryentry{strain_q}
{
  name=$\delta\epsilon$,
  description={Effective strain}
}

@@

\newglossaryentry{Ceps}
{
  name=C$_{\epsilon}$,
  description={Normalizing constant}
}

\newglossaryentry{eps-neighborhood}
{
 name=$\epsilon$,
  description={Neighbourhood size}
}

\newglossaryentry{heps}
{
  name=h$_{\epsilon}$,
  description={Band width}
}

@@

\newglossaryentry{Wn}
{
  name=$w_{n}$,
  description={Current crack opening}
}

@@

\newglossaryentry{gamma-lme}
{
  name=$\widehat{\gamma}$,
  description={Dimensionless parameter to control the value of the
    thermalization parameter \gls{LME-beta} of the \acrshort{lme} shape functions.}
}

@@

\journal{Engineering Fracture Mechanics}

@@

\textcolor{blue}{The \acrshort{mpm}~\cite{Sulsky1994} is a meshfree Lagrangian-Eulerian method where the continuum \gls{domain} is discretised with a finite set of material points (also known as particles in this research) $\hat{\Omega} = \{ \vec{x}_p, p \in \mathcal{C} \} \subset \Omega$, where $\mathcal{C} = 1, \ldots, n_p$. Any particle field such as position, velocity, mass, stain and stress denoted by \gls{x}, \gls{v}, \gls{m}, \gls{strain} and \gls{stress}, respectively, are assigned to each material point. In the present research, any particle quantity is denoted by ($\square_p$). In addition to this set of material points, a set of fixed background nodes is introduced to compute the balance of momentum equation. This set of nodes is defined as  $\textbf{X} = \{ \vec{x}_I, I \in \mathcal{B} \} \subset \mathbb{R}^{2}$, where $\mathcal{B} = 1, \ldots, n_I$. In the present research, any nodal quantity is denoted by ($\square_I$).}

@@

function width is defined as $U(\vec{x}_p,N_I) \coloneqq \sum_I N_I |\vec{x}_p - \vec{x}_I |^2$, and \gls{LME-beta} is a regularisation o \textit{thermalization} parameter such that for

@@

(0,\infty)$. Notice that \gls{LME-beta} has units of $[L]^{-2}$,

@@

enough efficient method. For an uniform nodal spacing, \gls{LME-beta} can be

@@

\subsection{Fracture modelling approach}
\label{sec:2.3}
Within the \textcolor{red}{\acrshort{mpm} formulation context}, fracture can be
modelled by failing particles according to a suitable failure
criterion. When material points are failed, they are assumed to have
null stress tensor. To reproduce this behaviour in the present
research, the eigensoftening algorithm is introduced in the 
\acrshort{mpm} framework as an alternative approach to the decohesion
model \cite{Schreyer_2002}. The eigensoftening concept
was \textcolor{red}{developed initially} by Navas {\it et al.}
(2017)\cite{Navas_2017_ES} as an extension for quasi-brittle materials
of the eigenerosion proposed by Pandolfi \& Ortiz
(2012)\cite{Pandolfi_2012} for fracture of brittle materials. The
  evaluation of both approaches \cite{Navas_2017_ES} against
  experimental measures show that the eigenerosion algorithm significantly overestimates the tensile stress and the strain peaks, Meanwhile eigensoftening captures the forces and crack
patterns accurately. Furthermore, the accuracy of this algorithm has been proven for \textcolor{red}{mixed-mode} fracture patterns such as the ones obtained with \textcolor{red}{\acrfull{frc}}, \cite{Navas_2018_ES}.\\

The key idea behind the eigenerosion algorithm is the computation of the
energy-release rate attendant to the failure of material point $p$,
\begin{align}
  \label{eq:energy-release-EE}
&G_p^{k+1} = \frac{C_{\epsilon}}{m_p^{k+1}}  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q W_q^{k+1}\\
  \label{eq:mass-EE}
&m_p^{k+1} =  \sum_{x_q^{k+1} \in
  B_{\epsilon}(x_p^{k+1})} m_q  
\end{align}
where $B_{\epsilon}(x_p^{k+1})$ is a $^n$dimensional sphere of radius \gls{eps-neighborhood}
centered at $x_p^{k+1}$. The particles which lie under this sphere are
known as the \gls{eps-neighborhood}-neighborhood of the material point (see
\cite{Pandolfi_2012}). This concept is conveniently sketched in
Figure \ref{fig:Failed-particles}.
%%%%%%%%%%%
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{Figure-Particle-failed}
  \caption{Scheme of a linear cohesive law, where the shaded area is
    \gls{Gf}, \gls{ft} is the tensile strength, and \gls{Wc} is the critical
    opening displacement.}
  \label{fig:Failed-particles}
\end{figure}
%%%%%%%%%%%
Other parameters are the mass of the neighborhood $m_p^{k+1}$, the
current free-energy density per unit mass  $W_q^{k+1}$ and the
normalizing constant \gls{Ceps}, which also defines the \gls{eps-neighborhood}-neighborhood size, being $\epsilon = C_\epsilon h$, and $h$ is a measure of the distance between nodes.

The failure criterion \textcolor{blue}{considers that a material point fails}
when $G_p^{k+1}$ surpasses the critical energy release rate that
measures the material-specific energy, \gls{Gf}. The convergence of this
approach has been analyzed by Schmidt {\it et al.}
(2009)\cite{Schmidt_2009}, who proved that \gls{Gf} converges to the Griffith
fracture when the discretisation size tends to zero. It is necessary to
point out that, when a material point overpass the critical energy, its
contribution to the internal forces vector is set to zero. \textcolor{red}{However,} its
contribution to the mass matrix is preserved.\\

@@

the static fracture energy per unit of area, \gls{Gf}. Notice how a
cohesive crack appears when the maximum tensile strength, \gls{ft}, is
reached. Once the crack, \gls{Wn}, reaches the value of the critical
crack opening, \gls{Wc}, a stress-free crack is attained.

@@

    \gls{Gf}, \gls{ft} is the tensile strength, and \gls{Wc} is the critical

@@

\gls{eps-neighborhood}-neighborhood of the material point $\vec{x}_p^{k+1}$ as

@@

material point in the \gls{eps-neighborhood}-neighborhood. At this point, an \textit{effective} strain \gls{strain_q} is introduced in order to obtain the variation of

@@

\delta\epsilon_q$. Now, with the assumption that the effective

@@

  \frac{\delta\tens{\epsilon}_p}{m_p} \sum_{x_q^{k+1} \in

@@

 \tens{\sigma}_{\epsilon,p}\ \delta\epsilon_p$. The softening behaviour is

@@

tensile strength, \gls{ft}. This consists of a reduction of the internal

@@

\cite{Navas_2017_ES} introduced a band width parameter \gls{heps} in order to relate the crack opening with the strain.

@@

opening \gls{Wn} within the band width \gls{heps}. Therefore, 

@@

\begin{table}
  \centering
  \begin{tabular}[]{c c c c c c c}
    \hline
      &   $\rho$   & $\text{f}_c$ & $\text{f}_t$ & \gls{Gf} &   E   & $\text{d}_{max}$ \\
      & (kg/m$^3$) &     (MPa)    &     (MPa)    & (N/m) & (GPa) & (mm) \\
    \hline
Value &    2368    &     102.7    &      5.4     &  141  &   31  &  12 \\
    \hline
  \end{tabular}
  \caption[Mechanical properties of thje concrete]{Mechanical
    properties of the high strenght concrete.}
  \label{tab:concrete-properties}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For the present research, a 2D setting of the experiment has been
adopted. Both the hammer and the concrete beam are explicitly
represented in order to properly capture the impact. Several levels of
discretisation were required to assess the objectiveness of the
obtained results. The results here exposed are from a discretisation
of 21576 material points and 3943 nodes. To achieve better results, an
unstructured grid layout has been adopted focusing the minimum nodal
spacing, 1.2 mm, in the middle of the beam and in the impact surface
of the hammer. For this simulation, \gls{gamma-lme} was fixed to
1.8, this low value of the regularization parameter smooth high stress
overshots after the impact. Here the stabilization properties of the \acrshort{lme}
described in \cite{Molinos2020} becomes crucial to achieve
non-spurious results. Standard linear element were tested
with unsatisfactory results since grid-crossing occurs in the hammer
head producing non realistic forces.\\

@@

\begin{algorithm}[H]
  \DontPrintSemicolon
  \KwInput{For each $p$, $\epsilon$-neighbourhood, $f_{t,p}$, $h_{\epsilon,p}$, \gls{Wc}}
  \KwOutput{Return damage parameter $\chi = \{\chi_p\}$}
  \KwData{Testing set $x$}
  $\chi_p \leftarrow \chi_p^{k}$ \;
  \For{$p$ to $N_p$}
  {
    \If{$\chi_p = 0$ \textbf{and} $\epsilon_{f,p} = 0$}
    {
      \For{$q \in B_{\epsilon,p}$}
      {
        $\sigma_{q,I} \leftarrow \text{getEigenvaluesOf} (\sigma_{q}) $ \;
        \If{$\chi_q < 1$}
        {
          $ \sum m_p\sigma_{p,I} \leftarrow \sum m_p\sigma_{p,I} + m_q\sigma_{q,I}$ \;
        }
        $ m_p \leftarrow m_p + m_q$ \;
      }
      $\sigma_{p,\epsilon} \leftarrow \frac{1}{m_p} \sum m_p\sigma_{p,I}$\;
      \If{$\sigma_{p,\epsilon} > f_{t,p}$}
      {
        $\epsilon_{f,p} = \epsilon_{I,p}$ \;  
      }
      \ElseIf{$\chi_p \neq 1$  \textbf{and} $\epsilon_{f,p} > 0$}
      {
        $\chi_p^{k+1} \leftarrow \min\Big \{1 , \max \{\chi_p^{k},
        \frac{(\epsilon_{I,p}- \epsilon_{f,p})\ h_{\epsilon,p}}{w_c}
        \} \Big \}$ \;
      }
    }
  }
    \label{alg-eigens}
    \caption{Compute damage parameter $\chi$}
\end{algorithm}
